---
title: 'Project 1: The Class'
output: html_document
---

```{r warning=FALSE, echo=FALSE, results='hide', message=FALSE}
# Data cleaning
# Load libraries
library(xlsx)
library(reshape)
library(stringr)
library(ggplot2)
library(cowplot)
library(devtools)
library(Rcpp)
library(httr)
library(rCharts)
library(rjson)
library(vcd)
library(ellipse)

# Read xlsx file
raw.data <- read.xlsx("../data/Survey_Response.xlsx", sheetName = "Form Responses 1")
data <- raw.data

# Drop empty columns
data <- data[, colSums(is.na(data)) < nrow(data)]


# Clean up variable names
cleaned.names <- c('waitlist', 'program','tools','exp.R.manipulation',
          'gender','text.editor','exp.R.graphics','exp.R.advanced','exp.R.reproducible','exp.Matlab','exp.Github')
names(data) <- cleaned.names

# Split tools column into multiple columns
## Count the maximum number of commas per row in the tools column
num.commas <- max(sapply(gregexpr(",", data$tools, fixed = TRUE), function(x) max(length(x)))) + 1
## Split the tools column in separate columns
tool.cols <- colsplit(data$tools,"[,]",names=sapply(1:num.commas, function(x) paste('tools', x, sep='')))
tool.cols <- sapply(tool.cols, str_trim)
unique.tool.values <- unique(unlist(lapply(tool.cols, FUN=unique)))
tool.matrix <- matrix(0, nrow = nrow(tool.cols),20)
tool.names <- unique.tool.values
colnames(tool.matrix) <- tool.names
tool.matrix <- as.data.frame(tool.matrix)
for (i in 1:ncol(tool.cols) ) {
      for (tool.name in tool.names) {
          tool.matrix[which(tool.cols[, i] == tool.name), tool.name] <- 1
          }
}
## Add cleaner tool names
clean.tool.names <- c("R", "Excel", "Matlab", "RStudio", "Github", "SQL", "Shell", "Python", "SPSS", "ggplot2", "GDrive", "Web", "C", "Dropbox", "Regex", "Stata", "LaTeX", "Sweave", "XML", "Lattice")
names(tool.matrix) <- clean.tool.names
data <- cbind(data, tool.matrix)
data <- subset(data, select=-c(tools))

# Clean up inconsistent program values
levels(data$program)
sort(table(data$program))
levels(data$program)[levels(data$program)=="Ms in ds"] <- "IDSE (master)"
levels(data$program)[levels(data$program)=="MSDS"] <- "IDSE (master)"
levels(data$program)[levels(data$program)=="PhD Biomedical Informatics"] <- "Ph.D."
levels(data$program)[levels(data$program)=="Data Science"] <- "Unknown"

# Clean up text editor values
levels(data$text.editor)
sort(table(data$text.editor))
# Remove extra spaces
data$text.editor <- factor(str_trim(data$text.editor))
# TextWrangler
levels(data$text.editor)[levels(data$text.editor)=="textwrangler"] <- "TextWrangler"
levels(data$text.editor)[levels(data$text.editor)=="Text Wrangler"] <- "TextWrangler"
levels(data$text.editor)[levels(data$text.editor)=="textWrangler"] <- "TextWrangler"
# Sublime Text
levels(data$text.editor)[levels(data$text.editor)=="Sublime"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="Sublime Text 2"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="sublime"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="Sublime Text!"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="sublime text 2"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="Sublime 2"] <- "Sublime Text"
levels(data$text.editor)[levels(data$text.editor)=="sublime text"] <- "Sublime Text"
# Jupyter
levels(data$text.editor)[levels(data$text.editor)=="I used jupyter last semester"] <- "Jupyter"
## Ipython is now known as Jupyter
levels(data$text.editor)[levels(data$text.editor)=="Ipython"] <- "Jupyter"
# None
levels(data$text.editor)[levels(data$text.editor)=="haven't used any"] <- "None"

# Assign level ordering to experience variables
experience.cols <- names(data)[grepl("^exp.", names(data))]
for (exp.col in experience.cols) {
  data[, exp.col] <- factor(data[, exp.col], levels=levels(data[, exp.col])[c(4, 1, 2, 3)], ordered=TRUE)
}
```


# Background
The following analysis examines the 114 survey responses from our EDAV class.

# Waitlist
- 16% of the class started on the waitlist.
```{r echo=FALSE}
barplot(prop.table(table(data$waitlist)), las=1, main='Percent of class on waitlist', cex.names=.8, horiz=TRUE, xlab='Percent of class')
```

# Gender
- 71% of the class is male, 28% female, and 1% says their gender doesn't matter.
```{r echo=FALSE}
barplot(prop.table(table(data$gender)), las=1, main='Gender distribution', cex.names=.5, horiz=TRUE, xlab='Percent of class')
```

# Program
- Looking at the distribution of students by program we see that almost 50% of our students are Master in data-science, and 20% are in certificate program.
  
```{r results="asis", echo=FALSE }
# Donut with Programs
pie.program <- rCharts$new()
pie.program <- nPlot(~program, data = data, type = 'pieChart')
pie.program$chart(donut = TRUE)
pie.program$print('iframe',  include_assets=TRUE)
```

# Tools
- R, Python, and Excel are the most commonly used tools among students.
```{r echo=FALSE}
# Total numer of use with all tools.
barplot(sort(colSums(data[, 11:30]) / nrow(data), decreasing = FALSE), las = 1, cex.names = 0.9, xlab = "Percent of students", main = "Experience with tools", hori=TRUE)
```

## Hierarchical clustering reveals distinct clusters of tool usage
```{r echo=FALSE}
corr.matrix <- cor(data[, 11:30])
distance <- dist(corr.matrix)
plot(hclust(distance), 
     main="Hierarchical clustering of tools", xlab="Tool")
```

### We interpret these clusters as:
- Excel & productivity tools
- Engineering tools
- Miscellaneous languages
- R and R packages

#### A correlation plot reveals the granular relationships between tools:
- There are some competitive products that are positively correlated while others are negatively correlated. For instance: dropbox and google drive are positively correlated while R and Python are negatively correlated. Perhaps it's because dropbox and google drive have similar usability while R and Python are less interchangeable.

```{r echo=FALSE}
ordering <- hclust(distance)$labels[hclust(distance)$order]
color.map <- colorRamp(c("#CC0000","white","#3366CC"), space="Lab")
plotcorr(corr.matrix[ordering,ordering], 
         col=rgb(color.map((corr.matrix[ordering,ordering]+1)/2), maxColorValue=255),
         mar = c(0.1, 0.1, 0.1, 0.1), type='lower')
text(5.5, 19, labels = 'Excel & productivity', col = rgb(0,0,0,.8))
text(10, 15, labels = 'Engineering', col = rgb(0,0,0,.8))
text(14, 9, labels = 'Misc.', col = rgb(0,0,0,.8))
text(18, 5, labels = 'R', col = rgb(0,0,0,.8))
```

# Text editors
- The distribution of text editor shows that RStudio is heavily favored.
```{r echo=FALSE}
# find out top 7 code editor tools used most.
by_preferred_tool <- sort(table(data$text.editor), decreasing = TRUE)[1:7]
by_preferred_tool['Other'] <- sum(sort(table(data$text.editor), decreasing = TRUE)[8:18])
barplot(by_preferred_tool / sum(by_preferred_tool), las = 1, cex.names = 0.6, xlab = "Percent", main = "Percent of users by text editor", horiz=TRUE)
```

```{r echo=FALSE, results='asis'}
#Sankey Graph
data_t=as.data.frame(table(data$program,data$text.editor))
colnames(data_t) <- c("source", "target", "value")
data_t=data_t[data_t[,3]>0,]
#remove all data with 0 relations
sankeyPlot <- rCharts$new()
sankeyPlot$setTemplate(script = "./rCharts_d3_sankey-gh-pages/layouts/chart.html")
sankeyPlot$setLib('http://timelyportfolio.github.io/rCharts_d3_sankey/libraries/widgets/d3_sankey')
sankeyPlot$set(
  data = data_t,
  nodeWidth = 15,
  nodePadding = 10,
  layout = 32,
  width = 750,
  height = 500,
  units = "TWh",
  title = "Editor-Program"
)

sankeyPlot$save('sankey.html',  cdn = TRUE)
```
- The sankey chart shows that the preference for RStudio is explained by the large number of statistic and data-science masters students.
- We can see text editor preferences by program: few students of statistics prefer different editors while 40% of data-science students do not use RStudio. 

<iframe width="900" height="520" frameborder="0" src = 'sankey.html' ></iframe>

# Experience by tool
- When it comes to R, students are more experienced with manipulation than with other skills.
- Students have little to no experience with most of the tools below


``````{r results="asis", echo=FALSE }

col.names.data <- c("exp.R.advanced","exp.Github","exp.R.reproducible"
                 ,"exp.R.manipulation","exp.R.graphics","exp.Matlab")
matrix={}
for (name.col in col.names.data){
  row=cbind(name.col,as.character(data[,name.col]))
  matrix=rbind(matrix,row)
}
data.boxplot=as.data.frame(matrix)
colnames(data.boxplot)=c("tool","experience")
data.boxplot$experience <- factor(data.boxplot$experience, c("None","A little","Confident","Expert"))

levels(data.boxplot$tool)[levels(data.boxplot$tool)=="exp.R.advanced"] <- "R.advanced"
levels(data.boxplot$tool)[levels(data.boxplot$tool)=="exp.R.reproducible"] <- "R.reproducible"
levels(data.boxplot$tool)[levels(data.boxplot$tool)=="exp.R.manipulation"] <- "R.manipulation"
levels(data.boxplot$tool)[levels(data.boxplot$tool)=="exp.R.graphics"] <- "R.graphics"
levels(data.boxplot$tool)[levels(data.boxplot$tool)=="exp.Matlab"] <- "Matlab"
levels(data.boxplot$tool)[levels(data.boxplot$tool)=="exp.Github"] <- "Github"


n2<-rCharts$new()
n2 = nPlot( ~tool,group="experience", data = data.boxplot, type = "multiBarChart")
n2$chart(stacked=TRUE)
n2$print('inline',  include_assets=TRUE)

```



# Tool expertise by program
-	Data Science Certification students tend to have the highest proportion of students with no experience with each tool, but they are pretty good at using Github.
-	Most other programs are familiar with skills in R.
-	Among all these six tools, experts are minority. Statistics major has the highest proportion of experts.
```{r echo=FALSE, warning=FALSE}
programs <- c('IDSE (master)', 'Data Science Certification', 'Statistics (master)', 'Other masters')
par(mfrow=c(3,2))

exp.by.program <- data[data$program %in% programs,]
exp.by.program['program'] <- factor(exp.by.program$program, levels=unique(exp.by.program$program))
barplot(prop.table(table(exp.by.program$exp.R.manipulation, exp.by.program$program), 2), las=1, main='R Manipulation', cex.names=.5)
barplot(prop.table(table(exp.by.program$exp.R.graphics, exp.by.program$program), 2), las=1, main='R Graphics', cex.names=.5)
barplot(prop.table(table(exp.by.program$exp.R.advanced, exp.by.program$program), 2), las=1, main='R Advanced', cex.names=.5)
barplot(prop.table(table(exp.by.program$exp.R.reproducible, exp.by.program$program), 2), las=1, main='R reproducible', cex.names=.5)
barplot(prop.table(table(exp.by.program$exp.Matlab, exp.by.program$program), 2), las=1, main='Matlab', cex.names=.5)
barplot(prop.table(table(exp.by.program$exp.Github, exp.by.program$program), 2), las=1, main='Github', cex.names=.5, legend.text = TRUE, 
        args.legend = list(x = "topright", bty = "n", inset=c(-.12, -1.3)))
```





``````{r results="asis", echo=FALSE, message=FALSE }
col.names.data=c("exp.R.advanced","exp.Github","exp.R.reproducible"
                 ,"exp.R.manipulation","exp.R.graphics","exp.Matlab")
matrix={}
for (name.col in col.names.data){
  row=cbind(name.col,cbind(as.character(data[,"program"]),as.character(data[,name.col])))
  matrix=rbind(matrix,row)
}
data.splot.tool=as.data.frame(matrix)
colnames(data.splot.tool)=c("tool","program","use")
levels(data.splot.tool$tool)[levels(data.splot.tool$tool)=="exp.R.advanced"] <- "R.advanced"
levels(data.splot.tool$tool)[levels(data.splot.tool$tool)=="exp.R.reproducible"] <- "R.reproducible"
levels(data.splot.tool$tool)[levels(data.splot.tool$tool)=="exp.R.manipulation"] <- "R.manipulation"
levels(data.splot.tool$tool)[levels(data.splot.tool$tool)=="exp.R.graphics"] <- "R.graphics"
levels(data.splot.tool$tool)[levels(data.splot.tool$tool)=="exp.Matlab"] <- "Matlab"
levels(data.splot.tool$tool)[levels(data.splot.tool$tool)=="exp.Github"] <- "Github"

data.splot.f=data.splot.tool[data.splot.tool$use != "None",]
data.final=as.data.frame(table(data.splot.f$program,data.splot.f$tool))
data.s=as.data.frame(table(data.splot.tool$tool,data.splot.tool$use))
colnames(data.s) <- c("source", "target", "value")
colnames(data.final) <- c("source", "target", "value")
data_sfinal=rbind(data.s,data.final)

#remove all data with 0 relations
sankeyPlotTool <- rCharts$new()
sankeyPlotTool$setTemplate(script = "./rCharts_d3_sankey-gh-pages/layouts/chart.html")
sankeyPlotTool$setLib('http://timelyportfolio.github.io/rCharts_d3_sankey/libraries/widgets/d3_sankey')
sankeyPlotTool$set(
  data = data_sfinal,
  nodeWidth = 15,
  nodePadding = 10,
  layout = 35,
  width = 750,
  height = 500,
  units = "TWh",
  title = "Editor-Program"
)

sankeyPlotTool$save('sankey_tool.html',  standalone = TRUE)


```
This Chart shows a relation between *program* , *tool* and *experience level*. We removed the association between program and tool when the level of experience is None. In Master Data-Science (IDSE) Github and Matlab are the most unknown topic and in certification program have a poor experience with R.reproducible. However 30% of all students do not have any knowledge in Matlab and 91% of them have worked with R.Manipulation.
<iframe width="900" height="520" frameborder="0" src = 'sankey_tool.html' ></iframe>